---
title: "MOFA Cross-Validation Benchmarking"
author: Mike Love, Wancen Mu
date: "`r format(Sys.Date(), '%d %b %Y')`"
params:
  which_py: '/usr/bin/python'
  mofa_maxiter: 500
  savedata: true
  rerun: false
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MOFA Cross-Validation Benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

To DO

* python deps
* docker

----------

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, cache = TRUE,  collapse = TRUE, comment = "#>" )
```

# Setup

## load required packages

```{r, eval=TRUE, message=FALSE}
library(data.table)
library(purrr)
library(Seurat)
library(ggplot2)
library(gridExtra)
library(caret)
library(magrittr)
library(matrixStats)
library(MOFA2)
library(dplyr)
library(GGally)
```

## parameters

```{r}
mofa_maxiter <- params$mofa_maxiter
which_py <- params$which_py
savedata <- params$savedata
```

## set up python

```{r, eval=FALSE, include=FALSE}
if (!reticulate::py_module_available('mofapy2')) {
  # specify python version
  which_py <- '/usr/bin/python'
  which_py <- which_py
  reticulate::use_python(which_py, required = T)
}
```

## set up output directory
```{r}
## directory to save the outputs
if (!dir.exists('savedata')) {
  dir.exists('savedata')
}
```

## run MOFA

Name the views for convenience
```{r, eval=TRUE}
views<-c("RNA expression","Enhancer accessibility", "Promoter accessibility","Enhancer methylation", "Promoter methylation" )
views_nick<-c("RNA","enh_atac","pro_atac","enh_methy","pro_methy")
nfactor<-5
```


### load pre-saved MOFA object from publication

Below is the code used for running the cross-validation, which takes approximately 2 hours. This code chunk is unevaluated, and instead we load the saved objects in the previous code chunk.
```{r}
# load full data model as vignette describes
# Original full model has 11 factors, we extract data and reconstruct using 5 factors indicated lineage
load(url("ftp://ftp.ebi.ac.uk/pub/databases/mofa/scnmt_gastrulation/gastrulation_scnmt_mofa.RData"))
```

### cross-validation runs
```{r}
sample_metadata <- MOFAmodel@samples_metadata
# Construct data for model input
data<-list()
for (q in views) {
  data[[q]]<-cbind(MOFAmodel@data[[q]]$E5.5,MOFAmodel@data[[q]]$E6.5,MOFAmodel@data[[q]]$E7.5)
}
## imputation #########################
### featureWise Mean ###################
featureWiseMean <- function(m) t(apply(m, 1, function(x) ifelse(is.na(x), mean(x,na.rm=TRUE), x)))
impute_featMean<-list()
for(p in views){
  impute_featMean[[p]]<-featureWiseMean(data[[p]])
}
## 4 fold cross validation #####################
# To keep get same value, foldMem were saved in the mofacv.Rdata and are already loaded 
##########################################################################################
kFold=4
trainIndex=groupKFold(colnames(data$`RNA expression`),k=kFold)
foldMem=rep(0,ncol(data$`RNA expression`))
for(i in 1:kFold){
  foldMem=foldMem+i*!(1:ncol(data$`RNA expression`) %in% trainIndex[[i]])
}
```

```{r, eval=params$rerun}
## cross validation model ##########################
for(i in 1:kFold){
  # a is index for trainging data
  a=which(foldMem!=i)
  dat<-list("RNA expression"=data$`RNA expression`[,a],"Enhancer accessibility"=data$`Enhancer accessibility`[,a],"Promoter accessibility"=data$`Promoter accessibility`[,a],
           "Enhancer methylation"=data$`Enhancer methylation`[,a], "Promoter methylation"=data$`Promoter methylation`[,a])

MOFAmodel_train <- create_mofa(dat,groups =sample_metadata$group[a])
# Model options
model_opts <- get_default_model_options(MOFAmodel_train)
model_opts$num_factors<-nfactor
# Training options
train_opts <- get_default_training_options(MOFAmodel_train)
train_opts$maxiter<-mofa_maxiter
data_opts<-get_default_data_options(MOFAmodel_train)
MOFAmodel_train <- prepare_mofa(
  object = MOFAmodel_train,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
MOFAmodel_train <- run_mofa(MOFAmodel_train,outfile = sprintf("savedata/model_%s.hdf5",i))
}
```

```{r}
## Construct contribution of each view from cv model #######################
cvContribution_RNA_Mat=list()
cvContribution_enh_atac_Mat=list()
cvContribution_pro_atac_Mat=list()
cvContribution_enh_methy_Mat=list()
cvContribution_pro_methy_Mat=list()
for (i in 1:4) {
MOFAmodel_train<-load_model(sprintf("savedata/model_%s.hdf5",i))
a=which(foldMem!=i)
weight_RNA_train<-(get_weights(MOFAmodel_train,views = "RNA expression"))$"RNA expression"
weight_enh_atac_train<-(get_weights(MOFAmodel_train,views = "Enhancer accessibility"))$"Enhancer accessibility"
weight_pro_atac_train<-(get_weights(MOFAmodel_train,views = "Promoter accessibility"))$"Promoter accessibility"
weight_enh_methy_train<-(get_weights(MOFAmodel_train,views = "Enhancer methylation"))$"Enhancer methylation"
weight_pro_methy_train<-(get_weights(MOFAmodel_train,views = "Promoter methylation"))$"Promoter methylation"

cvContribution_RNA=t(impute_featMean$`RNA expression`[,-a])%*%weight_RNA_train
cvContribution_enh_atac=t(impute_featMean$`Enhancer accessibility`[,-a])%*%weight_enh_atac_train
cvContribution_pro_atac=t(impute_featMean$`Promoter accessibility`[,-a])%*%weight_pro_atac_train
cvContribution_enh_methy=t(impute_featMean$`Enhancer methylation`[,-a])%*%weight_enh_methy_train
cvContribution_pro_methy=t(impute_featMean$`Promoter methylation`[,-a])%*%weight_pro_methy_train

cvContribution_RNA_Mat[[i]]=cvContribution_RNA
cvContribution_enh_atac_Mat[[i]]=cvContribution_enh_atac
cvContribution_pro_atac_Mat[[i]]=cvContribution_pro_atac
cvContribution_enh_methy_Mat[[i]]=cvContribution_enh_methy
cvContribution_pro_methy_Mat[[i]]=cvContribution_pro_methy
}
```

```{r, eval=params$rerun}
#### Full model ###################
data2<-list("RNA expression"=data$`RNA expression`,"Enhancer accessibility"=data$`Enhancer accessibility`,"Promoter accessibility"=data$`Promoter accessibility`,
            "Enhancer methylation"=data$`Enhancer methylation`, "Promoter methylation"=data$`Promoter methylation`)

MOFAmodel <- create_mofa(data2,groups =sample_metadata$group)
# Model options
model_opts <- get_default_model_options(MOFAmodel)
model_opts$num_factors<-nfactor
# Training options
train_opts <- get_default_training_options(MOFAmodel)
train_opts$maxiter<-mofa_maxiter
data_opts<-get_default_data_options(MOFAmodel)
MOFAmodel <- prepare_mofa(
  object = MOFAmodel,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

MOFAmodel <- run_mofa(MOFAmodel,outfile = "savedata/model_full.hdf5")
```


```{r}
MOFAmodel <- load_model("savedata/model_full.hdf5")
weight_full<-list()
fullContribution_featMean<-list()
for(p in views){
  weight_full[[p]]<-(get_weights(MOFAmodel,views = p))[[p]]
}
for(p in views){
  fullContribution_featMean[[p]]<-t(impute_featMean[[p]])%*%weight_full[[p]]
}
fullContribution2<- do.call("cbind", fullContribution_featMean)
colnames(fullContribution2)<-paste(rep(views_nick, times = 1, each = 5),c("fac1","fac2","fac3","fac4","fac5"),sep = ".")
fullContribution2<-scale(fullContribution2)
fullContribution2<-data.table(fullContribution2,sample=rownames(fullContribution2))
fullContribution2=fullContribution2%>% left_join(sample_metadata,by="sample")
## Store cvContribution for each cross validation ##########
cvContribution<-list()
for(i in 1:4){
  a<-cbind(cvContribution_RNA_Mat[[i]],
                               cvContribution_enh_atac_Mat[[i]],
                               cvContribution_pro_atac_Mat[[i]],
                               cvContribution_enh_methy_Mat[[i]],
                               cvContribution_pro_methy_Mat[[i]])
  colnames(a)<-paste(rep(views_nick, times = 1, each = 5),c("fac1","fac2","fac3","fac4","fac5"),sep = ".")
  a<-data.table(a,sample=rownames(a))
cvContribution[[i]]<-left_join(a,sample_metadata,by="sample")
}

save(cvContribution_RNA_Mat,
     cvContribution_enh_atac_Mat,
     cvContribution_pro_atac_Mat,
     cvContribution_enh_methy_Mat,
     cvContribution_pro_methy_Mat,foldMem,fullContribution2,cvContribution, file = "savedata/mofacv.RData")
```

## Analysis using loaded cross validation result ######################
```{r, eval=TRUE}
load("savedata/mofacv.RData")
##### Check factors order #################
a<-data.frame(fullContribution2[,1:5],"cell"=row.names(fullContribution2))
for (i in 1:4) {
  a<-merge(a,data.frame(cvContribution[[i]][,1:5],"cell"=rownames(cvContribution_RNA_Mat[[i]])),by="cell",all=T)
  # colnames(cvContribution_RNA_Mat[[i]])<-c("Factor1", "Factor2", "Factor3", "Factor4")
}
rna<-a[2:length(a)]
colnames(rna)<-paste(rep(c("full","cv1","cv2","cv3","cv4"), times = 1, each = 5),c("fac1","fac2","fac3","fac4","fac5"),sep = ".")

## save each factor location of all views into a list ################## 
fac.loc<-list()
for (i in 1:nfactor) {
  fac.loc[[i]]<-seq(i,dim(fullContribution2)[2],by=length(views))
}
cvContribution2<-cvContribution

###### change the order according to RNA correlation#########
for (i in 1:4) {
  mat <- cor(rna[,c(1:5,(5*i+1):(5*i+5))],use="pairwise.complete.obs")[1:5,6:10]
  # derive highest correlation among cv factors for each full data factor
  rm <- rowMaxs(abs(mat))
  o <- order(rm, decreasing=TRUE)
  matches <- numeric(5)
  for (row in o) {
    # save correct cv data factor order into matches vector
    matches[row] <- which.max(abs(mat[row,]))
    if(max(mat[row,])!=max(abs(mat[row,]))){
      # if direction is false, change contribution into the opposite direction
      cvContribution2[[i]][,fac.loc[[row]]]<--1*as.data.frame(cvContribution[[i]])[,fac.loc[[matches[row]]]]
    }else{
      cvContribution2[[i]][,fac.loc[[row]]]<-as.data.frame(cvContribution[[i]])[,fac.loc[[matches[row]]]]
    }
    # correlations of matched factors become 0 to avoid affect next unmatched factors 
    mat[,matches[row]] <- 0
  }
}

##### combind all the cv list object into a full dataframe #############
cvContribution_df<- do.call("rbind", cvContribution2)
cvContribution_value<-cvContribution_df[,1:25]
cvContribution_value<-scale(cvContribution_value)
cvContribution_df[,1:25]<-as.data.frame(cvContribution_value)

#########################################################
# Contribution plot ###################
## Full dataset ##############
## General ############

# pdf(file="savedata/full2.pdf", width=6, height=6)
for (i in 1:5) {
  g<-ggcorr(fullContribution2[,fac.loc[[i]]],hjust = 0.75, size = 4, color = "grey50",label = TRUE, label_size = 3, label_round = 2, label_alpha = TRUE)
  print(g)
}
# dev.off()

## Cross validation dataset ###################
## General #################
# pdf(file="savedata/cv.pdf", width=6, height=6)
for (i in 1:5) {
  g<-ggcorr(as.data.frame(cvContribution_df)[,fac.loc[[i]]],hjust = 0.75, size = 4, color = "grey50",label = TRUE, label_size = 3, label_round = 2, label_alpha = TRUE)
  print(g)
}
# dev.off()
```


```{r}
sessionInfo()
```
